A Classical Density-Functional Theory for Describing Water Interfaces

### 2013-08-12 papers/water-saft/paper.tex
Merged and added concatination of energy files in correct order for water. Added data and small changes to paper for commit Tested everything on node09.

Water, the universal solvent, is of critical practical importance, and
a continuum description of water is in high demand for a solvation
model.  A number of recent attempts to develop improved solvation
models for water have built on the approach of classical
density-functional theory (DFT)~\cite{jeanmairet2013molecular,
  zhao2011molecular, zhao2011new, ramirez2005direct,
  ramirez2005density, levesque2012solvation, levesque2012scalar}.
Classical DFT is based on a description of a fluid written as a free
energy functional of the density distribution.  There are two
approaches used to construct a classical DFT for water.  The first is
to choose a functional form that is fit to properties of the bulk
liquid at a given temperature and pressure
\cite{jeanmairet2013molecular} ~\fixme{cite Jeanmairet and similar}.
Using this approach, it is possible to construct a functional that
reproduces the exact second-order response function of the liquid
under the fitted conditions.  This class of functional, however, will
be less accurate at other temperatures or pressures---and in the
inhomogeneous scenarios in which solvation models are applied.  The
second approach is construct a functional by applying liquid-state
theory to a model system, and then fit the model to experimental data
such as the equation of state.

\fixme{stuff about water}.  An important aspects of water
is its ability to form hydrogen bonds.  Each water molecule contains
two hydrogen bond donor sites and two acceptor sites.
%
One promising approach for an efficient continuum description of water
is that of classical density-functional theory (DFT), which is an
approach for evaluating the free energy and thermally averaged density
of fluids in an arbitrary external potential~\cite{ebner1976density}.
The foundation of classical DFT is the Mermin
theorem\cite{mermin1965thermal}, which extends the Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} to non-zero temperature.
Classical DFT is a natural framework for creating a more flexible
theory of hydrophobicity that can readily describe interaction of
water with arbitrary external potentials---such as potentials
describing strong interactions with solutes or surfaces.

A widely used family of models used in the development of classical
density functionals is based on Statistical Associating Fluid Theory
(SAFT).

SAFT is a theory for fluids in which hydrogen bonding plays a significant
role\cite{chapman1989saft, muller2001molecular}. SAFT is used to
accurately model the equations of state of both pure fluids and
mixtures over a wide range of temperatures and pressures.
\fixme{Talk about and cite papers using SAFT to build DFTs.}

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/energy-vs-diameter}
\end{center}
\caption{ Excess chemical potential per area versus radius for a single
hydrophobic rod immersed in water. This should have an asymptote equal
to the surface tension at room temperature, and it agrees well with
the surface tension in Figure~\ref{fig:surface-tension}. The inset for
rods with a very small radius shows the linear relationship expected based on
Equation~\ref{contactvaluethm}.}
\label{fig:energy-vs-diameter}
\end{figure}

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
In Figure~\ref{fig:energy-vs-diameter} we show the excess chemical
potential at room temperature, scaled by the solvent accessible
surface area of the hard rod, plotted as a function of hard-rod
radius.  We define the hard-rods radius as the radius from which water
is excluded.  For rods with radius larger than 0.5~nm or so, this
reaches a maximum value of 75~mN/m, which is slightly higher than
macroscopic surface tension.  In the limit of very large rods, this
value will decrease and approach the macroscopic value.  As seen in
the inset of Fig.~\ref{fig:energy-vs-diameter}, for rods with very
small radius (less than about 0.5~\AA) the excess chemical potential
is proportional to volume, satisfying Equation~\ref{contactvaluethm},
which results from the contact-value theorem.

We show in Figure~\ref{fig:density-single-rod} density profiles for
different radii rods, as well as the prediction for the contact value
of the density as a function of rod radius, as computed from the
excess chemical potentials plotted in
Figure~\ref{fig:energy-vs-diameter}.  The agreement between these
curves confirms that our functional satisfies the contact-value
theorem and that our minimization is well converged.  As expected, as
the radius of the rods becomes zero the contact density approaches the
bulk density, and as the radius becomes large, the contact density
will approach the vapor density.

>>>>

Water, the universal solvent, is of critical practical importance, and
a continuum description of water is in high demand for a solvation
model.  A number of recent attempts to develop improved solvation
models for water have built on the approach of classical
density-functional theory (DFT)~\cite{jeanmairet2013molecular,
  zhao2011molecular, zhao2011new, ramirez2005direct,
  ramirez2005density, levesque2012solvation, levesque2012scalar}.
Classical DFT is based on a description of a fluid written as a free
energy functional of the density distribution.  There are two
approaches used to construct a classical DFT for water.  The first is
to choose a functional form that is fit to properties of the bulk
liquid at a given temperature and pressure
\cite{jeanmairet2013molecular} ~\fixme{cite Jeanmairet and similar}.
Using this approach, it is possible to construct a functional that
reproduces the exact second-order response function of the liquid
under the fitted conditions.  This class of functional, however, will
be less accurate at other temperatures or pressures---and in the
inhomogeneous scenarios in which solvation models are applied.  The
second approach is construct a functional by applying liquid-state
theory to a model system, and then fit the model to experimental data
such as the equation of state.  Classical DFT is a natural framework
for creating a more flexible theory of hydrophobicity that can readily
describe interaction of water with arbitrary external
potentials---such as potentials describing strong interactions with
solutes or surfaces.

An important aspects of water
is its ability to form hydrogen bonds.  Each water molecule contains
two hydrogen bond donor sites and two acceptor sites.

A widely used family of models used in the development of classical
density functionals is based on Statistical Associating Fluid Theory
(SAFT). SAFT is a theory for fluids in which hydrogen bonding plays a
significant role\cite{chapman1989saft, muller2001molecular}. SAFT is
used to accurately model the equations of state of both pure fluids
and mixtures over a wide range of temperatures and pressures.
\fixme{Talk about and cite papers using SAFT to build DFTs.}

@@

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
In Figure\fixme{deleted} we show the excess chemical
potential at room temperature, scaled by the solvent accessible
surface area of the hard rod, plotted as a function of hard-rod
radius.  We define the hard-rods radius as the radius from which water
is excluded.  For rods with radius larger than 0.5~nm or so, this
reaches a maximum value of 75~mN/m, which is slightly higher than
macroscopic surface tension.  In the limit of very large rods, this
value will decrease and approach the macroscopic value.  As seen in
the inset of Fig.~\fixme{deleted}, for rods with very
small radius (less than about 0.5~\AA) the excess chemical potential
is proportional to volume, satisfying Equation~\ref{contactvaluethm},
which results from the contact-value theorem.

We show in Figure~\ref{fig:density-single-rod} density profiles for
different radii rods, as well as the prediction for the contact value
of the density as a function of rod radius, as computed from the
excess chemical potentials plotted in
Figure~\fixme{deleted}.  The agreement between these
curves confirms that our functional satisfies the contact-value
theorem and that our minimization is well converged.  As expected, as
the radius of the rods becomes zero the contact density approaches the
bulk density, and as the radius becomes large, the contact density
will approach the vapor density.
